---
title: "Review Lab 1"
author: "Moshe Weiss"
output: pdf_document
---

These are exercises to practice for the exam. You do not need to hand this lab in as an assignment.

* Create a vector of 1000 draws from a normal distribution centered at 50 with variance 25.

```{r}
norms = rnorm(1000,50,25)
```

* Round those distributions down to the closest integer.

```{r}
floor(norms)
```

* Round those distributions up to the closest integer.

```{r}
ceiling(norms)
```

* Create a vector that looks like 1,1,1,1,1,2,2,2,2,2, ..., 100,100,100,100,100. Do it in as few operations as possible.

```{r}
repVec = c(rep(1:100,each =5))
repVec
```

Use the `which` function to find the indices of numbers that are divisible by 7.

```{r}
which(repVec%%7==0)
```


* Create a nominal factor vector that looks like A,A,A,B,B,B,C,C,C,D,D,D,D, ..., Z,Z,Z,Z and make the reference category Z.

```{r}
nomVec = rep(0, length(LETTERS)-1,each=3)
names(nomVec) = rep(LETTERS[1:length(LETTERS)-1],each=3)
nomVec
```

* Create a list mapping "A" => 1, "B" => 2, ..., "Z" => 26

```{r}
x = list()
x[LETTERS] = (1:26)
x
```

* Use `lapply` to sum over the values in the list.

```{r}
#nope
```


* Create a matrix of size 2 x 26 where the first row is the list's keys and the second row is the list's values.

```{r}
matrix(names(x),x)
rbind(names(x),(1:26))
```

* Write a function that runs a function f (assumed to return a scalar) on each column of a matrix and aggregates the results in a vector.

```{r}
X = matrix()
```

Load the boston housing data frame.

```{r}
data(Boston, package="MASS")
```

Create a list object that holds 10 dataframes where each consists of the boston housing data frame whose observations are sampled with replacement.

```{r}
listFrames =list()
for(i in 1:10){
  listFrames[[i]] = Boston[sample(1:length(Boston),replace = TRUE), ]
}
listFrames
```


Cross-tab the variable chas with rad.


```{r}
table(Boston$chas,Boston$rad)
```

Create a new variable in the data frame called `fake_y` equal to `3 * age + 6 * lstat - 100 * crim  +` iid random realizations from $N(0,1)$.

```{r}
Boston$fake_y=3*Boston$age + 6*Boston$lstat - 100*Boston$crim + rnorm(nrow(Boston),0,1)
summary(Boston$fake_y)
```

Find the deciles of `fake_y` i.e. the min, the 10th percentile, the 20th percentile, ..., the 90th percentile and the max.

```{r}
?quantile
quantile(Boston$fake_y,seq(0,1,.1))

```


Find the subset of the data frame where `fake_y` is negative.

```{r}

```


Create a new variable in the data frame which binarizes `fake_y` by being 1 if positive and 0 if negative.

```{r}

```
Fit an OLS model of 'fake_y` using `age` and find the estimates of the intercept and slope and RMSE and R^2.

```{r}
ourOLS = lm(fake_y~age,Boston)
coef(ourOLS)
summary(ourOLS)$residuals
summary(ourOLS)$r.squared
```











































